---
title: "GDAPC_capstone project"
author: "Ogunbiyi Tolulope Ayobami"
date: "2022-08-02"
output: html_document
---

## Setting up my environment
Notes: loading of package 'tidyverse' and 'cyclistic_bike_share' dataset needed for this analysis

```{r loading packages}
library(tidyverse)
library(tidyr)
library(readr)
library(ggplot2)
library(lubridate)
```

## Loading the datasets
Notes: the 'cyclistic-share-bike'dataset is loaded in separate dataframes from january to december

```{r cyclistic_share_bike datasets}

jan <-read_csv("DAC8_case_study/202101-divvy-tripdata.csv")
feb <-read_csv("DAC8_case_study/202102-divvy-tripdata.csv")
mar <-read_csv("DAC8_case_study/202103-divvy-tripdata.csv")
apr <-read_csv("DAC8_case_study/202104-divvy-tripdata.csv")
may <-read_csv("DAC8_case_study/202105-divvy-tripdata.csv")
jun <-read_csv("DAC8_case_study/202106-divvy-tripdata.csv")
july <-read_csv("DAC8_case_study/202107-divvy-tripdata.csv")
aug <-read_csv("DAC8_case_study/202108-divvy-tripdata.csv")
sept <-read_csv("DAC8_case_study/202109-divvy-tripdata.csv")
oct <-read_csv("DAC8_case_study/202110-divvy-tripdata.csv")
nov <-read_csv("DAC8_case_study/202111-divvy-tripdata.csv")
dec <-read_csv("DAC8_case_study/202112-divvy-tripdata.csv")
```
## Executing basic data inspection and cleaning
Notes: checking out the overview of the dataset 

```{r Overview of the dataset}
glimpse(jan)
glimpse(feb)
glimpse(mar)
glimpse(apr)
glimpse(may)
glimpse(jun)
glimpse(july)
glimpse(aug)
glimpse(sept)
glimpse(oct)
glimpse(nov)
glimpse(dec)

```
## Apending the different dataframes into one
Notes: dataset are appended into one dataframe for easy readability, naming conventions and for further cleaning to achieve good analysis

```{r binding dataset into a dataframe}
ride_data <- bind_rows(jan, feb, mar, apr, may, jun, july, aug, sept, oct, nov, dec)
```

## Basic cleaning of the new dataframe
Notes: Overview using glimpse, colnames, head, and str

```{r overview of dataframe}
head(ride_data)
glimpse(ride_data)
str(ride_data)
colnames(ride_data)
```

## Further cleaning
Notes: cleaning to ensure data integrity
```{r remove duplicates}
ride_data_new <- distinct(ride_data)
View(ride_data_new)
```

## Further cleaning
Notes: cleaning to ensure data integrity
```{r remove null values and rename variables}
ride_data_new <- drop_na(ride_data)
ride_data_new <- ride_data_new %>% 
  rename(bike_type = rideable_type, user_type = member_casual)
```

## Insight into calculating the ride length
```{r calculate trip duration}
ride_data_cleaned <- mutate(ride_data_new, trip_duration = as.numeric(ended_at - started_at)/60)
```

## Insight into calculating weeks and month of ride occurence
```{r Calculate week and month trip occured}
ride_data_cleaned <- mutate(ride_data_cleaned, 
   trip_day = format(as.Date(ride_data_new$started_at), "%A"), 
   trip_month = format(as.Date(ride_data_cleaned$started_at), "%B"))
```

## Overview of the cleaned dataset
Notes: a preview of the data after it has been cleaned
```{r data preview}
glimpse(ride_data_cleaned)
```

## A dive into the cyclistic bike share data for insights
Notes: Insight into Average ride length of users
```{r calculate ave_ride_lenght Vs. Users_type }
avg_trip_duration <- ride_data_cleaned %>%
  group_by(user_type) %>%
  summarise(avg_trip_duration = mean(trip_duration))
```
## Visualise the average ride lenght of Users
```{r ave_ride_length Vs. User_type}
ggplot(avg_trip_duration)+
  geom_col(mapping=aes(x = user_type, y = avg_trip_duration, fill = user_type))+
  labs(title = "Average Trip Duration Vs. User_type", subtitle = "Cyclistic Customer")
```

## Average trip length by users on days of the week
```{r avg_ride_length by users Vs.days_of_week}
avg_trip_week <- ride_data_cleaned %>%
  group_by (trip_day, user_type) %>%
  summarise(number_of_trips = n(), average_trip_length = mean(trip_duration),
            .groups = 'drop') %>%
arrange(user_type, trip_day)
```
```{r visualisation of average_ride_length of users Vs. days_of_week}
ggplot(avg_trip_week)+
  geom_col(mapping = aes(x = trip_day, y = average_trip_length, fill = user_type))+
  labs(title = "Average Weekly Trip Vs. User_type", subtitle = "Cyclistic Customer")
```

## Number of trips by users on weekly basis
```{r count_of_rides by users Vs. days_of_the_week}
avg_trip_week <- ride_data_cleaned %>%
  group_by(trip_day, user_type) %>%
  summarise(number_of_trips = n(), average_trip_length = mean(trip_duration),
            .groups = 'drop') %>%
  arrange(user_type, trip_day)
```
```{r visualise count_of_rides by users Vs. days_of_the_week}
ggplot(avg_trip_week)+
  geom_col(mapping = aes(x = trip_day, y = number_of_trips, fill = user_type))+
  labs(title = "Number of Weekly Trip Vs. User_type", subtitle = "Cyclistic Customer") 
```


## Number of trips by users on monthly basis
```{r avg_ride_length by users Vs.month_of_year}

avg_trip_month <- ride_data_cleaned %>%
  group_by(trip_month, user_type) %>%
  summarise(number_of_trips = n(), average_trip_length = mean(trip_duration),
            .groups = 'drop') %>%
  arrange(user_type, trip_month) 
```
```{r visualisation of month_of_year by users Vs. average_ride_length}
ggplot(avg_trip_month)+
  geom_col(mapping = aes(x = trip_month, y = number_of_trips, fill = user_type))+
  labs(title = "Average Monthly Trip Vs. User_type", subtitle = "Cyclistic Customer")
```

